---
title: "01 — Download American Stories data"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = TRUE)
PROJECT_ROOT <- Sys.getenv("CAPLAN_ROOT")
if (PROJECT_ROOT == "") PROJECT_ROOT <- rprojroot::find_root(rprojroot::has_file("_config.R"))
knitr::opts_knit$set(root.dir = PROJECT_ROOT)
source(file.path(PROJECT_ROOT, "_config.R"))
```

Downloads the American Stories corpus year-by-year via the Python `datasets` library
(system Python, to avoid reticulate venv issues) and writes one Parquet per year to
`data_parquet/articles_<year>.parquet`. See `download_data.py`.

Start with the **primary analysis window** (`r min(YEARS_ANALYSIS)`–`r max(YEARS_ANALYSIS)`)
where the county shock datasets exist; download the full 1774–1960 span later for the
descriptive national-trend figures.

```{r download}
# Primary window first. For the full span call with no args (slow, large).
start_year <- as.integer(min(YEARS_ANALYSIS))
end_year   <- as.integer(max(YEARS_ANALYSIS))

message("Downloading American Stories ", start_year, "-", end_year, " ...")
exit_code <- system2(
  PYTHON_EXE,
  args = c(shQuote(file.path(PROJECT_ROOT, "download_data.py")),
           start_year, end_year),
  stdout = "", stderr = "", wait = TRUE
)
if (exit_code != 0) warning("download_data.py exited with code ", exit_code)
```

## Summary

```{r summary}
parquet_files <- list.files(DATA_PARQUET, pattern = "^articles_\\d{4}\\.parquet$",
                            full.names = TRUE)
message("Total per-year Parquet files: ", length(parquet_files))
if (length(parquet_files)) {
  info <- file.info(parquet_files)
  info$year <- as.integer(gsub(".*articles_(\\d{4})\\.parquet", "\\1", rownames(info)))
  message("Years: ", min(info$year), "-", max(info$year),
          " (", nrow(info), " files)")
  message("Total size: ", round(sum(info$size) / 1e9, 2), " GB")
}
```
