---
title: "03 — Link newspapers to publication county (FIPS)"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = TRUE)
PROJECT_ROOT <- Sys.getenv("CAPLAN_ROOT")
if (PROJECT_ROOT == "") PROJECT_ROOT <- rprojroot::find_root(rprojroot::has_file("_config.R"))
knitr::opts_knit$set(root.dir = PROJECT_ROOT)
source(file.path(PROJECT_ROOT, "_config.R"))
source(file.path(PROJECT_ROOT, "R", "helpers.R"))
library(arrow); library(data.table); library(dplyr)
```

Builds the crosswalk `data_panels/xwalk_lccn_county.parquet` mapping each newspaper
(LCCN / title) to its **publication county FIPS** — the join key for every shock dataset.
See docs/DATA_SOURCES.md §B1. Strategy: pull the Chronicling America title authority list
(LCCN → place of publication), join to the corpus, resolve place → county FIPS.

## 1. Get the Chronicling America title list

```{r ca-titles}
library(httr2); library(jsonlite)
ca_raw <- file.path(DATA_RAW, "chronicling_america")
dir.create(ca_raw, showWarnings = FALSE, recursive = TRUE)
titles_path <- file.path(ca_raw, "ca_titles.parquet")

if (!file.exists(titles_path)) {
  # The loc.gov / Chronicling America API lists newspapers with lccn, title, state,
  # county, city, place_of_publication. Paginate through the JSON feed.
  # TODO(human): confirm the endpoint/params for your LoC snapshot; a manual CSV
  # export dropped at data_raw/chronicling_america/ca_titles.csv also works.
  base <- "https://chroniclingamerica.loc.gov/newspapers/?format=json"
  resp <- request(base) |> req_user_agent("caplan-newspaper-bias/research") |>
    req_perform()
  js <- resp_body_json(resp, simplifyVector = TRUE)
  titles <- as.data.table(js$newspapers)   # columns include lccn, title, state, county, city
  arrow::write_parquet(titles, titles_path)
} else {
  titles <- as.data.table(read_parquet(titles_path))
}
message("CA titles: ", nrow(titles), " rows; cols: ",
        paste(names(titles), collapse = ", "))
```

## 2. Extract an LCCN from the corpus

```{r corpus-lccn}
ds <- open_dataset(file.path(DATA_PARQUET, "articles_cleaned"))
# Pull the distinct newspaper identity columns present in the corpus.
id_cols <- intersect(c("article_id", "lccn", "newspaper_name"), names(ds$schema))
nl <- ds |> select(all_of(id_cols)) |> distinct() |> collect() |> as.data.table()

# American Stories often encodes the LCCN (pattern "sn" + 8 digits) inside the
# article_id or a dedicated field. Try to recover it.
if (!"lccn" %in% names(nl)) {
  nl[, lccn := stringi::stri_extract_first_regex(article_id, "sn\\d{8}")]
}
message("Distinct papers: ", nrow(nl),
        "; with LCCN: ", sum(!is.na(nl$lccn)))
```

## 3. Resolve place of publication → county FIPS

```{r to-fips}
# Join corpus papers to CA titles on LCCN (preferred) then normalized title.
norm <- function(x) tolower(trimws(gsub("[^a-z ]", "", tolower(x))))
titles[, title_norm := norm(title)]
if ("newspaper_name" %in% names(nl)) nl[, title_norm := norm(newspaper_name)]

xwalk <- merge(nl, titles[, .(lccn, title_norm, state, county, city)],
               by = "lccn", all.x = TRUE, suffixes = c("", "_ca"))
# fallback join on title for rows that missed on LCCN
miss <- is.na(xwalk$state) & "title_norm" %in% names(xwalk)
if (any(miss)) {
  fb <- merge(xwalk[miss, .(article_id, title_norm)],
              titles[, .(title_norm, state, county, city)],
              by = "title_norm", all.x = TRUE)
  xwalk[miss, c("state", "county", "city") := fb[, .(state, county, city)]]
}

# (state, county name) -> 5-digit FIPS via usmap. County-name spellings vary;
# usmap::fips() handles most. Unmatched go to a patch file.
get_fips <- function(state, county) {
  out <- rep(NA_character_, length(state))
  for (i in seq_along(state)) {
    if (is.na(state[i]) || is.na(county[i])) next
    f <- tryCatch(usmap::fips(state = state[i], county = county[i]),
                  error = function(e) NA_character_)
    out[i] <- f
  }
  out
}
xwalk[, fips := get_fips(state, county)]

matched   <- xwalk[!is.na(fips)]
unmatched <- xwalk[is.na(fips)]
message(sprintf("FIPS matched: %d / %d (%.1f%%)",
                nrow(matched), nrow(xwalk), 100 * nrow(matched) / nrow(xwalk)))

write_parquet(matched, file.path(DATA_PANELS, "xwalk_lccn_county.parquet"))
fwrite(unmatched, file.path(DATA_PANELS, "xwalk_unmatched.csv"))
# TODO(human): hand-patch xwalk_unmatched.csv (county spellings, defunct counties,
# multi-county metros), append to the parquet, and re-run downstream steps.
```

## 4. Coverage diagnostics

```{r coverage}
# Add Census region for stratified sampling (step 04) and regional trends.
region_of <- function(st) {
  ne <- c("CT","ME","MA","NH","RI","VT","NJ","NY","PA")
  mw <- c("IL","IN","MI","OH","WI","IA","KS","MN","MO","NE","ND","SD")
  so <- c("DE","FL","GA","MD","NC","SC","VA","WV","DC","AL","KY","MS","TN","AR","LA","OK","TX")
  we <- c("AZ","CO","ID","MT","NV","NM","UT","WY","AK","CA","HI","OR","WA")
  fifelse(st %in% ne, "Northeast",
    fifelse(st %in% mw, "Midwest",
      fifelse(st %in% so, "South",
        fifelse(st %in% we, "West", NA_character_))))
}
matched[, region := region_of(state)]
matched[, .N, by = region][order(-N)] |> print()
write_parquet(matched, file.path(DATA_PANELS, "xwalk_lccn_county.parquet"))
```
