---
title: "04 — Stratified sample for labeling"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = TRUE)
PROJECT_ROOT <- Sys.getenv("CAPLAN_ROOT")
if (PROJECT_ROOT == "") PROJECT_ROOT <- rprojroot::find_root(rprojroot::has_file("_config.R"))
knitr::opts_knit$set(root.dir = PROJECT_ROOT)
source(file.path(PROJECT_ROOT, "_config.R"))
source(file.path(PROJECT_ROOT, "R", "helpers.R"))
library(arrow); library(data.table); library(dplyr)
set.seed(SEED)
```

Builds `data_panels/label_queue.parquet`: the passages to label (LLM + human audit, step 05).
Two ingredients (docs/PROJECT_PLAN.md §6.1): a **stratified random base** (across decade ×
region, so the model sees ordinary text and the relevance gate is calibrated) plus
**oversampled positives** for each bias using the seed lexicons in `R/helpers.R`
(`BIAS_SEED_LEXICONS`) — lexicons are for *finding candidates only*, never the final measure.

## Parameters

```{r params}
N_BASE     <- 4000   # stratified random passages (relevance-gate + negatives)
N_POS_PER  <- 1500   # oversampled candidate-positive passages PER bias
# => ~4000 + 4*1500 = ~10k passages to weak-label; audit a stratified subset by hand.
```

## Load cleaned corpus + region

```{r load}
ds <- open_dataset(file.path(DATA_PARQUET, "articles_cleaned"))
xwalk <- read_parquet(file.path(DATA_PANELS, "xwalk_lccn_county.parquet")) |>
  as.data.table()
key_col <- if ("lccn" %in% names(xwalk)) "lccn" else "article_id"

# Restrict to the analysis window to keep the sample on-target.
yrs <- as.integer(YEARS_ANALYSIS)
base_tbl <- ds |>
  filter(year %in% yrs) |>
  select(article_id, article, year, any_of(c("lccn", "newspaper_name"))) |>
  collect() |> as.data.table()

base_tbl <- merge(base_tbl, xwalk[, .(key = get(key_col), region)],
                  by.x = if ("lccn" %in% names(base_tbl)) "lccn" else "article_id",
                  by.y = "key", all.x = TRUE)
base_tbl[is.na(region), region := "Unknown"]
base_tbl[, decade := (year %/% 10) * 10]
message("Articles in window: ", nrow(base_tbl))
```

## Stratified random base

```{r base-sample}
strata <- base_tbl[, .(n = .N), by = .(decade, region)]
strata[, take := pmax(1, round(N_BASE * n / sum(n)))]
base_sample <- base_tbl[, .SD[sample(.N, min(.N, strata[.(decade[1], region[1]),
                              on = .(decade, region), take]))],
                        by = .(decade, region)]
base_sample[, sample_reason := "base_stratified"]
message("Base sample: ", nrow(base_sample))
```

## Oversample candidate positives per bias

```{r oversample}
pos_list <- list()
for (b in names(BIAS_SEED_LEXICONS)) {
  base_tbl[, sc := lexicon_score(article, BIAS_SEED_LEXICONS[[b]])]
  cand <- base_tbl[sc > 0][order(-sc)]
  take <- head(cand, N_POS_PER)
  take[, sample_reason := paste0("seed_", b)]
  pos_list[[b]] <- take
  message(sprintf("  %-12s candidates>0: %d, taken: %d", b, nrow(cand), nrow(take)))
}
base_tbl[, sc := NULL]
pos_sample <- rbindlist(pos_list, fill = TRUE)
```

## Combine, dedupe, write queue

```{r write}
queue <- rbindlist(list(base_sample, pos_sample), fill = TRUE)
queue <- unique(queue, by = "article_id")           # an article enters once
setnames(queue, "article", "text", skip_absent = TRUE)
keep <- intersect(c("article_id", "text", "year", "region", "decade",
                    "lccn", "newspaper_name", "sample_reason"), names(queue))
queue <- queue[, ..keep]

write_parquet(queue, file.path(DATA_PANELS, "label_queue.parquet"))
message("label_queue.parquet: ", nrow(queue), " passages")
print(queue[, .N, by = sample_reason][order(-N)])
```

Next: `python/llm_label.py` weak-labels this queue; `rmd/05` audits + builds train/dev/test.
