---
title: "04 — Stratified sample for labeling"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = TRUE)
PROJECT_ROOT <- Sys.getenv("CAPLAN_ROOT")
if (PROJECT_ROOT == "") PROJECT_ROOT <- rprojroot::find_root(rprojroot::has_file("_config.R"))
knitr::opts_knit$set(root.dir = PROJECT_ROOT)
source(file.path(PROJECT_ROOT, "_config.R"))
source(file.path(PROJECT_ROOT, "R", "helpers.R"))
library(arrow); library(data.table); library(dplyr)
set.seed(SEED)
```

Builds `data_panels/label_queue.parquet`: the ~10k passages for the jury (step 05). Two
ingredients (docs/PROJECT_PLAN.md §6.1): a **stratified random base** (decade × region — calibrates
the relevance gate and supplies negatives) plus **oversampled candidate-positives** per bias.

MEMORY-SAFE + FULL-ERA: we sample ids from the **06a lexscored** parquets (ids + lexicon scores,
*no text*), then fetch article text for **only the ~10k selected rows** — so this never loads the
corpus text into memory, and it spans every era RoBERTa will score (override `CAPLAN_SAMPLE_YEARS`).
The lexicon is a *candidate finder only*, never the final measure. (Requires 06a lexscored output.)

## Parameters

```{r params}
N_BASE    <- 4000L    # stratified random passages (relevance gate + negatives)
N_POS_PER <- 1500L    # candidate-positive oversample PER bias (rare biases keep all they have)
BIAS      <- names(BIAS_SEED_LEXICONS)
lexcols   <- paste0(BIAS, "_lex")
```

## Gather ids from the lexscored parquets (no text loaded)

```{r gather-ids}
lex_dir <- file.path(DATA_PARQUET, "articles_lexscored")
stopifnot(dir.exists(lex_dir))
lex_files <- list.files(lex_dir, pattern = "lexscored_\\d{4}\\.parquet$", full.names = TRUE)
.sy <- Sys.getenv("CAPLAN_SAMPLE_YEARS")                       # default: ALL eras
if (nzchar(.sy)) {
  sel <- unlist(lapply(strsplit(.sy, ",")[[1]], function(p)
    if (grepl("-", p)) do.call(seq, as.list(as.integer(strsplit(p, "-")[[1]]))) else as.integer(p)))
  fy  <- as.integer(sub(".*lexscored_(\\d{4}).*", "\\1", lex_files))
  lex_files <- lex_files[fy %in% sel]
}
stopifnot(length(lex_files) > 0)
message("sampling ids from ", length(lex_files), " lexscored years (no text loaded)")

# Per-year: a small base pool (ids only) + all per-bias candidates (lex>0). Never reads text.
base_per_year <- ceiling((N_BASE * 4) / length(lex_files))     # generous pool, post-stratified below
base_pool <- vector("list", length(lex_files))
cand_pool <- setNames(vector("list", length(BIAS)), BIAS)
for (i in seq_along(lex_files)) {
  d <- as.data.table(read_parquet(lex_files[i], col_select = c("article_id", "year", "lccn", lexcols)))
  base_pool[[i]] <- d[sample(.N, min(.N, base_per_year)), .(article_id, year, lccn)]
  for (b in BIAS) {
    cb <- d[get(paste0(b, "_lex")) > 0, .(article_id, year, lccn, sc = get(paste0(b, "_lex")))]
    if (nrow(cb)) cand_pool[[b]][[length(cand_pool[[b]]) + 1L]] <- cb
  }
  rm(d)
}
```

## Select ids: stratified base + per-bias candidate positives

```{r select-ids}
xwalk <- as.data.table(read_parquet(file.path(DATA_PANELS, "xwalk_lccn_county.parquet")))
reg   <- unique(xwalk[, .(lccn, region)], by = "lccn")

base_dt <- rbindlist(base_pool, fill = TRUE)
base_dt <- merge(base_dt, reg, by = "lccn", all.x = TRUE)
base_dt[is.na(region), region := "Unknown"]
base_dt[, decade := (year %/% 10L) * 10L]
strata  <- base_dt[, .(n = .N), by = .(decade, region)]
strata[, take := pmax(1L, round(N_BASE * n / sum(n)))]
base_sel <- base_dt[, .SD[sample(.N, min(.N, strata[.(decade[1], region[1]),
                            on = .(decade, region), take]))], by = .(decade, region)]
base_sel[, sample_reason := "base_stratified"]

pos_parts <- list()
for (b in BIAS) {
  cb <- rbindlist(cand_pool[[b]], fill = TRUE)
  ntook <- 0L
  if (nrow(cb)) {
    setorder(cb, -sc)
    take <- head(cb, N_POS_PER); take[, sample_reason := paste0("seed_", b)]
    pos_parts[[b]] <- take[, .(article_id, year, lccn, sample_reason)]
    ntook <- nrow(take)
  }
  message(sprintf("  %-12s candidates>0: %d, taken: %d", b, nrow(cb), ntook))
}
sel <- unique(rbindlist(c(list(base_sel[, .(article_id, year, lccn, sample_reason)]), pos_parts),
                        fill = TRUE), by = "article_id")
message("distinct passages selected: ", nrow(sel))
```

## Fetch text for the selected ids only, then write the queue

```{r fetch-text-write}
ds  <- open_dataset(file.path(DATA_PARQUET, "articles_cleaned"))
ids <- sel$article_id
txt <- ds |> filter(article_id %in% ids) |>
  select(article_id, article, newspaper_name) |> collect() |> as.data.table()

queue <- merge(sel, txt, by = "article_id", all.x = TRUE)
queue <- merge(queue, reg, by = "lccn", all.x = TRUE); queue[is.na(region), region := "Unknown"]
queue[, decade := (year %/% 10L) * 10L]
setnames(queue, "article", "text")
queue <- queue[!is.na(text) & nchar(text) > 0]
keep  <- intersect(c("article_id", "text", "year", "region", "decade",
                     "lccn", "newspaper_name", "sample_reason"), names(queue))
queue <- queue[, ..keep]
write_parquet(queue, file.path(DATA_PANELS, "label_queue.parquet"))
message("label_queue.parquet: ", nrow(queue), " passages")
print(queue[, .N, by = sample_reason][order(-N)])
```

Next (the jury, attended/paid):
`python/label_jury.py run --queue data_panels/label_queue.parquet --outdir data_panels/label_out`
(competitors vote → merge → Claude adjudicates contested → `final.parquet`); then `rmd/05` builds train/dev/test.
