---
title: "05 — Jury-label and build train/dev/test"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = TRUE)
PROJECT_ROOT <- Sys.getenv("CAPLAN_ROOT")
if (PROJECT_ROOT == "") PROJECT_ROOT <- rprojroot::find_root(rprojroot::has_file("_config.R"))
knitr::opts_knit$set(root.dir = PROJECT_ROOT)
source(file.path(PROJECT_ROOT, "_config.R"))
source(file.path(PROJECT_ROOT, "R", "helpers.R"))
library(arrow); library(data.table); library(dplyr)
set.seed(SEED)
```

Turns the labeling queue into supervised splits: (1) the **multi-LLM jury** (`python/label_jury.py`)
labels every passage — competitors vote, Claude adjudicates *only* contested rows; (2) map stances to
binary targets; (3) split **train/dev/test by title + year** to prevent leakage. **Fully automated —
no human-in-the-loop**; validity is anchored by the codebook **gold/adversarial suite** (CODEBOOK §5).
The jury makes **paid API calls** — run `python/label_jury.py preflight` first to confirm keys/models.

## 1. Run the multi-LLM jury (competitors → merge → Claude adjudicates contested)

```{r jury}
queue   <- file.path(DATA_PANELS, "label_queue.parquet")
outdir  <- file.path(DATA_PANELS, "label_out")
final_p <- file.path(outdir, "final.parquet")
stopifnot(file.exists(queue))
# label_jury.py reads keys from CAPLAN_KEYS_FILE (default LLM/Keys.env.txt) -- never committed.
# Resumable: re-running skips providers/rows already done. Delete final.parquet to force a full redo.
if (!file.exists(final_p)) {
  exit <- system2(PYTHON_EXE, args = c(
    shQuote(file.path(PROJECT_ROOT, "python", "label_jury.py")), "run",
    "--queue", shQuote(queue), "--outdir", shQuote(outdir),
    "--id-col", "article_id", "--text-col", "text"),
    stdout = "", stderr = "", wait = TRUE)
  if (exit != 0) stop("label_jury.py run exited with ", exit)
}
lab <- as.data.table(read_parquet(final_p))
message(sprintf("jury labels: %d  |  contested: %d (%.1f%%)  |  high-confidence: %d",
                nrow(lab), sum(lab$contested, na.rm = TRUE),
                100 * mean(lab$contested, na.rm = TRUE), sum(lab$confidence == "high", na.rm = TRUE)))
```

## 2. Binary targets (only stance = "endorse" marks a bias present)

```{r targets}
q <- as.data.table(read_parquet(queue))
d <- merge(q, lab, by = "article_id", all.x = TRUE)
d[, relevance := fifelse(is.na(relevance), 0L, as.integer(relevance))]
for (b in c("anti_market", "anti_foreign", "make_work", "pessimistic")) {
  d[, (paste0("y_", b)) := fifelse(get(paste0(b, "_stance")) %in% "endorse", 1L, 0L)]
}
message("label base rates (mean):")
print(d[, lapply(.SD, function(x) round(mean(x, na.rm = TRUE), 4)),
        .SDcols = patterns("^y_|^relevance$")])
print(d[, .N, by = confidence])
```

## 3. Train/dev/test split (by title × 5-year block, no leakage)

```{r split}
grp_col <- if ("lccn" %in% names(d) && d[, sum(!is.na(lccn))] > 0L) "lccn" else "newspaper_name"
d[, grp := paste0(get(grp_col), "_", (year %/% 5L) * 5L)]   # title × 5-year block
g <- sample(unique(d$grp)); n <- length(g)
train_g <- g[seq_len(floor(.70 * n))]
dev_g   <- g[(floor(.70 * n) + 1L):floor(.85 * n)]
test_g  <- g[(floor(.85 * n) + 1L):n]
splits  <- list(train = d[grp %in% train_g], dev = d[grp %in% dev_g], test = d[grp %in% test_g])
keep <- c("article_id", "text", "relevance", "confidence",
          paste0("y_", c("anti_market", "anti_foreign", "make_work", "pessimistic")))
for (nm in names(splits)) {
  write_parquet(splits[[nm]][, intersect(keep, names(splits[[nm]])), with = FALSE],
                file.path(DATA_PANELS, paste0(nm, ".parquet")))
  message(nm, ": ", nrow(splits[[nm]]), " passages")
}
```

Validity is anchored by the codebook **gold/adversarial suite** (relevance + endorse precision/recall;
CODEBOOK §5) — no human audit, by design. Next: RoBERTa trains on `train`/`dev`, evaluated on `test`
+ the gold suite (`rmd/06`).
