---
title: "06b — Baseline: TF-IDF/SVD features (validated RF path)"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = TRUE)
PROJECT_ROOT <- Sys.getenv("CAPLAN_ROOT")
if (PROJECT_ROOT == "") PROJECT_ROOT <- rprojroot::find_root(rprojroot::has_file("_config.R"))
knitr::opts_knit$set(root.dir = PROJECT_ROOT)
source(file.path(PROJECT_ROOT, "_config.R"))
source(file.path(PROJECT_ROOT, "R", "helpers.R"))
library(arrow); library(data.table); library(quanteda); library(Matrix); library(irlba)
BIAS <- names(BIAS_SEED_LEXICONS)
set.seed(SEED)
```

Builds the Random-Forest feature space exactly as `american-stories-incivility`
(`04_feature_engineering`): TF-IDF → SVD(200) on a training sample, plus metadata and the
per-bias lexicon scores. Generalized to the four biases. Two differences, both
method-preserving: (1) we keep the document-term matrix **sparse** (no dense 300k×10k
blow-up) so it scales to this corpus; (2) we **save the fitted transform**
(vocab + idf + SVD center/rotation) so `06c` can score the full corpus with the same
features — completing the deployment the predecessor left as future work.

## Parameters

```{r params}
N_BASE    <- 100000L   # stratified random base (negatives + relevance signal)
N_POS_PER <- 30000L    # cap of lexicon-positive oversample PER bias (rare biases keep all)
N_SVD     <- 200L
TOP_FEATS <- 10000L
YEARS     <- as.integer(YEARS_ANALYSIS)   # 1895:1945
```

## Stratified sample across analysis-window years (text + metadata + lexicon scores)

```{r sample}
cleaned_dir <- file.path(DATA_PARQUET, "articles_cleaned")
lex_dir     <- file.path(DATA_PARQUET, "articles_lexscored")

yds <- list.files(cleaned_dir, pattern = "^year=\\d{4}$", full.names = TRUE)
yrs <- as.integer(sub(".*year=(\\d{4}).*", "\\1", yds))
keep <- yrs %in% YEARS & file.exists(file.path(lex_dir, paste0("lexscored_", yrs, ".parquet")))
yds <- yds[keep]; yrs <- yrs[keep]
stopifnot(length(yds) > 0)

# Oversample lexicon-positives so rare biases (anti_foreign/make_work) have enough RF
# training positives, alongside a stratified random base (negatives + relevance signal).
# Read each year, join lexicon scores, peel off per-bias positives + a random base.
cols <- c("article_id","newspaper_name","article","word_count","ocr_nonalpha",
          "headline","byline","page","year")
base_per_year <- ceiling(N_BASE / length(yds))
pos  <- setNames(lapply(BIAS, function(.) vector("list", length(yds))), BIAS)
base <- vector("list", length(yds))
for (i in seq_along(yds)) {
  yr <- yrs[i]
  pf <- list.files(yds[i], pattern = "\\.parquet$", full.names = TRUE)[1]
  dt <- as.data.table(arrow::read_parquet(pf, col_select = cols))
  lx <- as.data.table(arrow::read_parquet(
    file.path(lex_dir, paste0("lexscored_", yr, ".parquet")),
    col_select = c("article_id", paste0(BIAS, "_lex"))))
  dt <- merge(dt, lx, by = "article_id", all.x = TRUE)
  for (b in BIAS) pos[[b]][[i]] <- dt[get(paste0(b, "_lex")) > 0]
  base[[i]] <- dt[sample(.N, min(.N, base_per_year))]
  rm(dt, lx); gc()
}
parts <- base
for (b in BIAS) {
  pb <- rbindlist(pos[[b]], fill = TRUE)
  if (nrow(pb) > N_POS_PER) pb <- pb[sample(.N, N_POS_PER)]
  message(sprintf("  %-12s positives in sample: %d", b, nrow(pb)))
  parts[[length(parts) + 1L]] <- pb
}
train <- unique(rbindlist(parts, fill = TRUE), by = "article_id")
rm(pos, base, parts); gc()
message("training sample: ", nrow(train), " articles (random base + per-bias positive oversample)")
```

## Metadata features

```{r meta}
pg <- suppressWarnings(as.integer(gsub("[^0-9]", "", train$page)))
train[, front_page  := fifelse(!is.na(pg) & pg == 1L, 1L, 0L)]
train[, headline_len := nchar(headline)]
train[, has_byline   := as.integer(nchar(trimws(byline)) > 0)]
train[, ocr_quality  := 1 - ocr_nonalpha]
meta_cols <- c("front_page", "headline_len", "word_count", "has_byline", "ocr_quality")
```

## TF-IDF (sparse) → SVD(200), saving the transform

```{r tfidf-svd}
toks <- tokens(corpus(train$article), remove_punct = TRUE, remove_numbers = TRUE,
               remove_symbols = TRUE) |>
  tokens_tolower() |> tokens_remove(stopwords("en")) |> tokens_wordstem()
dfmat <- dfm(toks)
dfmat <- dfm_trim(dfmat, min_termfreq = 50, min_docfreq = 20)
dfmat <- dfm_select(dfmat, names(topfeatures(dfmat, TOP_FEATS)))
vocab <- colnames(dfmat)
message("vocab size: ", length(vocab))

m   <- as(dfmat, "CsparseMatrix")                       # docs × vocab, counts (sparse)
idf <- log10(nrow(m) / Matrix::colSums(m > 0))          # quanteda-style inverse df
tfidf <- m %*% Matrix::Diagonal(x = idf)                # sparse count*idf, no densify
rm(toks, dfmat, m); gc()

svd <- irlba::prcomp_irlba(tfidf, n = N_SVD, center = TRUE, scale. = FALSE)
svd_x <- as.data.table(svd$x); setnames(svd_x, paste0("svd_", seq_len(ncol(svd_x))))
message("SVD components: ", ncol(svd_x))

saveRDS(list(vocab = vocab, idf = idf, center = svd$center, rotation = svd$rotation,
             n_svd = N_SVD),
        file.path(MODELS_DIR, "rf_feature_transform.rds"))
```

## Assemble and save the training feature matrix

```{r assemble}
feat <- cbind(
  train[, c("article_id", "newspaper_name", "year", paste0(BIAS, "_lex")), with = FALSE],
  train[, ..meta_cols],
  svd_x
)
write_parquet(feat, file.path(DATA_PANELS, "rf_features.parquet"))
message("rf_features.parquet: ", nrow(feat), " rows x ", ncol(feat), " cols")
print(names(feat)[1:15])
```
