---
title: "06c — Baseline: Random Forest train + calibrate + score (validated RF path)"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = TRUE)
PROJECT_ROOT <- Sys.getenv("CAPLAN_ROOT")
if (PROJECT_ROOT == "") PROJECT_ROOT <- rprojroot::find_root(rprojroot::has_file("_config.R"))
knitr::opts_knit$set(root.dir = PROJECT_ROOT)
source(file.path(PROJECT_ROOT, "_config.R"))
source(file.path(PROJECT_ROOT, "R", "helpers.R"))
library(arrow); library(data.table); library(quanteda); library(Matrix); library(ranger); library(pROC)
BIAS <- names(BIAS_SEED_LEXICONS)
set.seed(SEED)
NTHREAD <- max(1L, parallel::detectCores() - 2L)
quanteda_options(threads = NTHREAD)
```

Trains one `ranger` Random Forest per bias on **lexicon-derived pseudo-labels** (top-decile
= positive, bottom-half = negative), with grouped-by-newspaper CV and isotonic calibration —
exactly the `american-stories-incivility` `05_random_forest` method, generalized to four
biases. Then **scores the full window corpus** by applying the saved TF-IDF→SVD transform
(`06b`) batch-by-batch — the deployment the predecessor left as future work. Uses all CPU
cores (`ranger num.threads`, `quanteda threads`).

## 1. Train per-bias RF (pseudo-labels + grouped CV + calibration)

```{r train}
feat <- as.data.table(read_parquet(file.path(DATA_PANELS, "rf_features.parquet")))
svd_cols  <- grep("^svd_", names(feat), value = TRUE)
meta_cols <- c("front_page", "headline_len", "word_count", "has_byline", "ocr_quality")
feature_cols <- c(svd_cols, meta_cols)
for (c in feature_cols) set(feat, which(is.na(feat[[c]])), c, 0)

models <- list(); calibs <- list(); cv_tab <- list()
for (b in BIAS) {
  lx  <- feat[[paste0(b, "_lex")]]
  # Lexicon scores are zero-inflated, so percentile cutoffs collapse (p90 = p50 = 0).
  # Distant supervision by presence/absence: any hit = positive; no hit = negative,
  # subsampling the large negative pool to ~5x positives for a trainable balance.
  pos <- which(lx > 0); neg <- which(lx == 0)
  if (length(pos) < 20L) { message("  ", b, ": only ", length(pos), " positives, skipping"); next }
  neg <- sample(neg, min(length(neg), 5L * length(pos)))
  d <- feat[c(pos, neg)]; d[, y := factor(c(rep(1L, length(pos)), rep(0L, length(neg))))]

  nps <- unique(d$newspaper_name)
  fmap <- data.table(newspaper_name = nps, fold = sample(rep(1:5, length.out = length(nps))))
  d <- merge(d, fmap, by = "newspaper_name")
  actual <- as.integer(as.character(d$y)); preds <- numeric(nrow(d))
  for (k in 1:5) {
    tr_i <- d$fold != k
    rf <- ranger(x = d[tr_i, ..feature_cols], y = d$y[tr_i], num.trees = 500,
                 mtry = floor(sqrt(length(feature_cols))), min.node.size = 10,
                 probability = TRUE, num.threads = NTHREAD, seed = SEED)
    preds[!tr_i] <- predict(rf, d[!tr_i, ..feature_cols], num.threads = NTHREAD)$predictions[, "1"]
  }
  auc   <- as.numeric(pROC::auc(pROC::roc(actual, preds, quiet = TRUE)))
  brier <- mean((preds - actual)^2)
  calibs[[b]] <- fit_isotonic(preds, actual)
  models[[b]] <- ranger(x = d[, ..feature_cols], y = d$y, num.trees = 500,
                        mtry = floor(sqrt(length(feature_cols))), min.node.size = 10,
                        probability = TRUE, num.threads = NTHREAD, seed = SEED)
  cv_tab[[b]] <- data.table(bias = b, n = nrow(d), pos = sum(actual), auc = auc, brier = brier)
  message(sprintf("  %-12s n=%d pos=%d AUC=%.3f Brier=%.3f", b, nrow(d), sum(actual), auc, brier))
}
saveRDS(list(models = models, calibs = calibs, feature_cols = feature_cols),
        file.path(MODELS_DIR, "rf_baseline_models.rds"))
print(rbindlist(cv_tab, fill = TRUE))
```

## 2. Score the window corpus (batched, fitted transform applied article-by-article)

```{r score}
tr <- readRDS(file.path(MODELS_DIR, "rf_feature_transform.rds"))
center_proj <- as.numeric(tr$center %*% tr$rotation)   # 1 x n_svd
cleaned_dir <- file.path(DATA_PARQUET, "articles_cleaned")
lex_dir     <- file.path(DATA_PARQUET, "articles_lexscored")
out_dir     <- file.path(DATA_PARQUET, "articles_scored_baseline")
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

# Years to score: window years that have lexicon scores; optional CAPLAN_SCORE_YEARS override.
avail <- as.integer(sub(".*lexscored_(\\d{4}).*", "\\1",
           list.files(lex_dir, pattern = "lexscored_\\d{4}\\.parquet$")))
score_years <- sort(intersect(avail, as.integer(YEARS_ANALYSIS)))
ov <- Sys.getenv("CAPLAN_SCORE_YEARS")
if (nzchar(ov)) {
  sel <- unlist(lapply(strsplit(ov, ",")[[1]], function(p)
    if (grepl("-", p)) do.call(seq, as.list(as.integer(strsplit(p, "-")[[1]]))) else as.integer(p)))
  score_years <- sort(intersect(score_years, sel))
}
message("scoring years: ", paste(score_years, collapse = ", "))
B <- 100000L
# Optional per-year row cap for fast smoke-testing (CAPLAN_SCORE_LIMIT); unset = score all.
score_limit <- suppressWarnings(as.integer(Sys.getenv("CAPLAN_SCORE_LIMIT")))

featurize <- function(text) {
  dm <- tokens(corpus(text), remove_punct = TRUE, remove_numbers = TRUE, remove_symbols = TRUE) |>
    tokens_tolower() |> tokens_remove(stopwords("en")) |> tokens_wordstem() |> dfm()
  dm <- dfm_match(dm, features = tr$vocab)               # align columns to training vocab
  tfidf <- as(dm, "CsparseMatrix") %*% Matrix::Diagonal(x = tr$idf)
  sweep(as.matrix(tfidf %*% tr$rotation), 2, center_proj, "-")  # = (tfidf - center) %*% rotation
}

for (yr in score_years) {
  out <- file.path(out_dir, paste0("scored_", yr, ".parquet"))
  if (file.exists(out)) { message(yr, ": exists"); next }
  pf <- list.files(file.path(cleaned_dir, paste0("year=", yr)), pattern = "\\.parquet$", full.names = TRUE)[1]
  dt <- as.data.table(read_parquet(pf, col_select = c(
    "article_id","newspaper_name","lccn","date","page","year","word_count","ocr_nonalpha","headline","byline","article")))
  lx <- as.data.table(read_parquet(file.path(lex_dir, paste0("lexscored_", yr, ".parquet")),
    col_select = c("article_id", paste0(BIAS, "_lex"))))
  dt <- merge(dt, lx, by = "article_id", all.x = TRUE)
  if (!is.na(score_limit) && nrow(dt) > score_limit) dt <- dt[sample(.N, score_limit)]
  pg <- suppressWarnings(as.integer(gsub("[^0-9]", "", dt$page)))
  dt[, front_page := fifelse(!is.na(pg) & pg == 1L, 1L, 0L)]
  dt[, headline_len := nchar(headline)]
  dt[, has_byline := as.integer(nchar(trimws(byline)) > 0)]
  dt[, ocr_quality := 1 - ocr_nonalpha]

  n <- nrow(dt); res <- vector("list", ceiling(n / B))
  for (i in seq_len(ceiling(n / B))) {
    rows <- ((i - 1) * B + 1):min(i * B, n)
    newd <- as.data.table(featurize(dt$article[rows])); setnames(newd, svd_cols)
    newd <- cbind(newd, dt[rows, ..meta_cols])
    ob <- dt[rows, .(article_id, newspaper_name, lccn, date, year)]
    for (b in BIAS) {
      if (is.null(models[[b]])) next
      pc <- calibs[[b]](predict(models[[b]], newd, num.threads = NTHREAD)$predictions[, "1"])
      ob[[paste0("p_", b, "_rf")]]      <- pc
      ob[[paste0("endorse_", b, "_rf")]] <- as.integer(pc >= 0.5)
      ob[[paste0(b, "_lex")]]            <- dt[[paste0(b, "_lex")]][rows]
    }
    res[[i]] <- ob
  }
  tmp <- paste0(out, ".tmp"); write_parquet(rbindlist(res, fill = TRUE), tmp); file.rename(tmp, out)
  message(sprintf("  %d: scored %d articles", yr, n)); rm(dt, lx, res); gc()
}
```
