---
title: "Project Estimates & Actuals — API Cost and Runtime"
author: "Caplan / Newspaper-Biases project"
date: "Initiated 2026-05-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

**Living document.** It records the *predictions* made at project start and the *actuals*
measured as each stage runs. Update the **Actual** columns and add a **Change log** entry
whenever a stage completes or an estimate is revised. Predictions below were made on
**2026-05-25**, before the full-corpus run, and are calibrated by smoke tests on the
1895–1899 slice (§4).

---

## 1. API cost

**Structural fact:** API cost is driven by the **~10k LLM labeling sample**, *not* the
corpus. The full corpus (hundreds of millions of articles) is scored by the **local
RoBERTa + lexicon/RF on the GPU/CPU — $0 in API calls**. Local Llama (Ollama) is also free.
So corpus size does not move the bill; only the bounded labeling sample does.

### Prediction — per full labeling pass (~10k passages)

| Provider | Role | Calls | Est. cost / pass |
|---|---|---:|---:|
| Gemini Flash | competitor | ~10k | ~$1 |
| GPT-4o-mini | competitor | ~10k | ~$1.5 |
| Grok (fast) | competitor | ~10k | ~$3–5 |
| Llama (local) | competitor | ~10k | $0 |
| Claude Sonnet | adjudicator (~25% contested) | ~2.5k | ~$12 |
| **One pass** | | | **~$18–25** |

**Project total estimate: ~$40–80** (expect 2–4 passes: codebook/prompt tuning, gold-suite
validity loop, final clean run). Dominated by Claude adjudication.

*Assumptions:* ~600 input + ~120 output tokens/call; ~25% of passages contested → adjudicated;
representative early-2026 pricing. Biggest uncertainties: Grok pricing, contested rate, # passes.

*Cost-reduction levers:* batch endpoints (~50% off cloud calls); Haiku instead of Sonnet for
adjudication (~3–4× cheaper); tighter contested threshold.

### Actual spend (update as incurred)

| Item | Date | Actual | Notes |
|---|---|---:|---|
| Gold-suite eval (×2 runs, 34 passages) | 2026-05-25 | ~$0.10–0.20 | preflight + 2 full jury runs on the validity set |
| Full labeling pass #1 | — | TBD | |
| Full labeling pass #2 | — | TBD | |
| **Running total** | | **~$0.20** | |

---

## 2. Runtime & compute (full analysis window, ~150–250M articles)

```{r download-elapsed, results='asis'}
log <- file.path(Sys.getenv("CAPLAN_DATA", "C:/Users/jdamm/Caplan"), "download.log")
if (file.exists(log)) {
  ll <- grep("^\\d{4}-\\d{2}-\\d{2}", readLines(log), value = TRUE)
  ts <- as.POSIXct(substr(ll, 1, 19), format = "%Y-%m-%d %H:%M:%S")
  cat(sprintf("*Auto-pulled:* download.log spans %s → %s (%.1f h wall-clock, incl. restarts/stalls).\n",
              format(min(ts, na.rm = TRUE)), format(max(ts, na.rm = TRUE)),
              as.numeric(difftime(max(ts, na.rm = TRUE), min(ts, na.rm = TRUE), units = "hours"))))
} else cat("*download.log not found.*\n")
```

### Prediction — by stage

| Stage | Type | Estimate | Actual | Notes |
|---|---|---|---|---|
| Corpus download | wait (network) | ~5 h | TBD | ~75–110 GB/hr observed |
| Ingest tar.gz → parquet | CPU (parallel) | ~1–2 h | TBD | |
| Clean (02) | CPU | ~2–4 h | **SERIAL ~14 h → only 1770–1869 (~100 yrs), never reached window — bottleneck. Parallelized 2026-05-26 (§4).** | ~25 min / 25.7M (slice) |
| Geolink LCCN→county (03) | network (cached) | ~1–2 h | TBD | bounded by # titles |
| Lexicon scoring (06a) | CPU (parallel) | ~1–3 h | TBD | 25.7M in minutes (slice) |
| RF features + train (06b/c) | CPU | ~0.5–1 h | TBD | minutes on slice |
| **RF corpus scoring (06c)** | **CPU (heavy)** | **~4–12 h** | TBD | initial est.; REVISED 2026-05-25 -> days–weeks (Revision §2a) |
| LLM jury labeling (~10k) | API + local GPU | ~2–6 h | TBD | throughput-bound |
| RoBERTa train | GPU | ~1–2 h | TBD | small labeled set |
| **RoBERTa corpus scoring** | **GPU (heavy)** | **~8–24 h** | TBD | initial est.; REVISED 2026-05-25 -> weeks–months (incivility evidence §3; Revision §2a) |
| Shock data prep (08) | human/network | hours–days | TBD | NHGIS/NOAA/IPUMS extracts |
| Panel + DiD + robustness (07/09/10) | CPU | ~4–8 h | TBD | initial est.; REVISED 2026-05-25 -> ~1–3 h (DiD is fast per prior runs) |
| Figures/tables (11) | CPU | minutes | TBD | |

### Headline — initial estimate (2026-05-25)

- **Pure machine compute: ~2–3 days**, dominated by the two full-corpus scoring passes.
- The RF pass is **CPU-bound** and the RoBERTa pass is **GPU-bound**, so they **run
  concurrently** → critical path ≈ max(RF, RoBERTa) ≈ **~½–1 day**, not the sum.
- **Compressed wall-clock with overlap: ~1.5–2.5 days** of the machine working.
- **Calendar incl. human/data waits: ~3–5 days** to final DiD tables (shock-data acquisition
  and codebook/jury validation iteration are the non-compute long poles; they overlap compute).

### Revision §2a — runtime (2026-05-25, after real-world evidence)

> NOTE: the initial headline above is **too optimistic by ~1–2 orders of magnitude** on the
> corpus-scoring stages. Decisive evidence: the incivility project ran **~3 months** of
> transformer scoring on an RTX 4000 Ada (full corpus) and is still ~6 years from done (§3).
> *The initial estimate is preserved above; this is an additive correction, not a replacement.*

- **Buildable/validatable part** (data prep, jury labeling, RF/RoBERTa *training*, panel/DiD):
  **~days to ~2 weeks**.
- **Full-corpus scoring (RoBERTa stance + RF): ~weeks-to-months** — the dominant, timeline-gating
  cost, even on the Blackwell; the "hand off to long-running agents" part the plan anticipated.
- **Total to final DiD: ~weeks to a few months**, gated almost entirely by scoring.
- **Levers** (may pull below incivility's ~3 months, but not to days): window to 1895–1945;
  staged relevance gate (stance heads only on the ~5–10% relevant); Blackwell 24 GB vs 4000 Ada
  12 GB; fp16/bf16 + ONNX/TensorRT + tuned batch size. Best case ~1–4 weeks scoring — pending a
  measured one-year benchmark (§2 caveats).

### Revision §2b — full-corpus scope re-estimate + session lessons (2026-05-26)

> SCOPE CHANGE (§4): measurement — clean + lexicon + RF/RoBERTa **scoring** — now spans the
> **whole corpus**, not the 1895–1945 window. The window survives only as the DiD treatment span
> (08/09); more pre-treatment years strengthen the DiD, so out-of-window scores are signal, not
> waste. *Original §2/§2a numbers preserved above; this is an additive correction.*

**Cost: UNCHANGED (~$40–80 total).** API cost is set by the bounded ~10k labeling sample, not
corpus size — local RF/RoBERTa scoring is $0. Widening the sample to span all eras keeps it ~10k
(stratified over more decades), so the bill doesn't move. Scope ↑ ⇒ runtime ↑, **cost flat.**

**Runtime — full-corpus vs. the windowed §2/§2a estimates:**

| Stage | Windowed (prior) | Full-corpus (new) | Basis |
|---|---|---|---|
| Clean (02) | ~2–4 h (est) | **~3–4 h ACTUAL**, 2 workers | measured this session; was the stalled bottleneck, now fixed (parallel + OOM-resilient) |
| Lexicon (06a) | ~1–3 h | ~2–4 h | CPU-parallel; ~linear in articles |
| Features + RF train (06b/c) | ~0.5–1 h | ~1–2 h | more training data |
| **RF corpus scoring (06c)** | days–weeks | **weeks** (1.53× measured) | CPU; full÷window article ratio |
| **RoBERTa scoring (06)** | weeks–months | **months — dominant long pole** (1.53× measured) | GPU; anchored to incivility (§3) |
| Panel + DiD (07/09) | ~1–3 h | ~1–3 h (unchanged) | small; consumes the scored panel |

**Headline (full-corpus):** total-to-final-DiD is gated almost entirely by **full-corpus RoBERTa
scoring → months-class** — the long-running hand-off the plan always anticipated. Window years
score first, so analysis-ready outcome data lands in the first portion of the run. The full÷window
multiplier is now **PINNED (2026-05-26, clean complete 180/180): 307.2M articles total; 200.4M in the
1895–1945 window = 65.2 % of the corpus ⇒ full÷window = 1.53×.** The window already holds the
peak-volume 1890s–1910s decades (1900s+1910s ≈ 130M articles), so the full-corpus scope premium is
only ~50 % — *less* than the 1.7–2.5× first guessed. Decade volumes (M articles): 1850s 11 · 1860s
14 · 1870s 18 · 1880s 26 · 1890s 46 · 1900s 61 · 1910s 69 · 1920s 28 · 1930s 10 · 1940s 9 · 1950s 6.

**What I've learned (this session):**

1. **The "cheap" stages can dominate.** Clean — penciled in at minutes–hours — became the
   bottleneck when serial (~14 h, never reached the window). Each big year peaks **~25–35 GB in
   RAM** (text normalization transiently doubles the article vector). Fix: small fixed worker pool
   + per-year `tryCatch` (one OOM no longer aborts the run) + window-first dispatch → ~3–4 h actual.
   *Profile memory and parallelize the unglamorous prep stages; don't eyeball them.*
2. **Headroom beats max parallelism.** A 10-worker clean OOM'd; contention with a parallel job made
   it worse. Now policy: **≤80 % RAM/CPU/GPU, ≥20 % free.** A killed worker or RAM fight costs more
   than the speed it chased.
3. **Scope discipline (the big one).** Measure everything; let only the small final DiD step apply
   the window. This raises scoring load (1.53× measured) but is methodologically right — and the cheapest
   stage must not constrain the most expensive ones.
4. **Cost is corpus-size-invariant** — reconfirmed. The scope change moves runtime, not the bill.
5. **Check the data before building external pipelines.** Geolink (03) needed *no* LoC API — the
   city/county/state is in the newspaper-title parenthetical (reused incivility's parser),
   sidestepping a Cloudflare bot-block entirely.
6. **"Reuse the validated repo" has limits.** Incivility used a newspaper-level *Coughlin-exposure*
   design with **no county geolink**, so Caplan's county-shock design is genuinely new there.
7. **Environment friction is budgetable.** The Store build of Dropbox crashes (root of the chronic
   sync problem; switched to the classic client) and Windows WMI calls hang after big memory frees
   (removed from unattended code). Small things, real hours.

### Caveats

- The two corpus-scoring estimates are **extrapolated from small slices** — the biggest
  uncertainty. A timed full-year benchmark (one real window year, end-to-end) would replace
  the extrapolation with measured numbers; **do this once the corpus finishes downloading.**
- Everything is **per-year checkpointed/resumable** — reboots/stalls cost restart time, not redo.
- If RF corpus scoring drags, narrow it to just the window years the DiD actually needs.

### Operational policy (from prior-run experience)

- **Resource-headroom policy — ≤80% utilization, ≥20% free (2026-05-26).** When launching parallel
  agents (clean workers, RF scoring, GPU RoBERTa workers), size them to use **at most ~80% of RAM,
  CPU, and GPU**, leaving **≥20% free** for the OS and any concurrent work on the box (notably the
  parallel `didgpu` DiD run). **Do not push for max speed** — a process restart or RAM fight costs
  far more than the speedup. Prompted by overnight #2, where a 10-worker clean OOM'd while another
  process held ~465 GB committed; `02` now sizes workers by 80% of *total* RAM (not momentarily-free)
  so the buffer survives mid-run contention. **Applies to every stage's launch math below.**
- **Checkpoint scoring frequently.** Scorers write **per batch** (atomic part files; resume by
  skipping completed batches), so a crash/reboot loses ≤ one batch (minutes), not a whole year.
  Implemented in `06c`; the RoBERTa scorer will match. Target: save ≤ every 30 min.
- **Parallel GPU scoring workers.** Shard by year (independent writes, no collisions), each
  checkpointing. Cap worker count by the **80% VRAM ceiling** (≤~19 GB of the 24 GB Blackwell):
  if each model instance needs ~5–6 GB, that is **~3 workers, not a blind 4** — benchmark N = 1/2/3.
  (Earlier note said ~4 on the Blackwell vs 2 on the 12 GB 4000 Ada; revised down to respect the
  headroom policy.) The **RF** scorer stays a **single process** but pinned to **~80% of cores**
  (`num.threads ≈ 19`, not all 24), since the box may be shared with the DiD work.
- **DiD is fast** (minutes–~1 h) — not a bottleneck; scoring dominates.
- **Years:** the manifest lists 180 per-year *files* (1770–1964, with calendar gaps); the count
  with *usable* data after cleaning is lower (~172, per prior project), and the DiD only needs the
  **1895–1945 window** (~50 yrs). Confirm the usable-year count after ingest+clean.

---

## 3. Calibration data points — smoke tests on the 1895–1899 slice (2026-05-25)

| Metric | Observed |
|---|---|
| Articles cleaned (1895–1899) | 25.77M (~82% kept after OCR/length filters) |
| Lexicon any-hit rate | anti_market 0.69% · pessimistic 1.13% · anti_foreign 0.04% · make_work 0.02% |
| RF baseline AUC (random-sample, thin) | am 0.90 · af 0.89 · mw 0.79 · pe 0.84 |
| RF positives after oversampling | am 30k (cap) · af 11.5k · mw 4.9k · pe 30k (cap) |
| RF baseline AUC (oversampled, final) | am 0.92 · af 0.95 · **mw 0.93 (was 0.79)** · pe 0.85 |
| Jury vs gold suite | relevance 1.0; endorse precision 1.0; recall 1.0 on 3/4 biases |
| Jury contested rate (gold suite) | 32.4% (11/34) — note: harder than a typical corpus sample |

### External real-world calibration (decisive for runtime)

The **incivility project** — same *American Stories* corpus, the predecessor pipeline — has
run on a Lenovo (i9 13th-gen, 128 GB RAM, **RTX 4000 Ada 12 GB**) for **~3 months**, largely
on the transformer scoring step, and still has **~6 years left to score** before analysis.
That implies real sustained throughput on the order of **tens of articles/sec** — roughly
1–2 orders of magnitude below the ~thousands/sec the original estimate assumed. **Anchor the
corpus-scoring runtime to this evidence, not to extrapolation.** Caplan's better GPU + windowed
scope + staged relevance gate may beat it, but the cost is the weeks-to-months class.

---

## 4. Change log

- **2026-05-25** — Initial cost and runtime predictions recorded (pre-full-run). Calibrated by
  the 1895–1899 smoke tests. RF baseline + multi-LLM jury validated end-to-end on the slice;
  full corpus still downloading (~137/180 at time of writing).
- **2026-05-25 (revision)** — Runtime estimate **substantially revised upward** for corpus
  scoring after real-world evidence: the incivility project has spent ~3 months on transformer
  scoring (RTX 4000 Ada, full corpus) and is not done (§3). Corpus scoring is now estimated at
  **weeks-to-months** (was hours) — the timeline-gating step and the long-running hand-off.
  Original optimistic estimate retained for the record. API-cost estimate unchanged (it's
  driven by the bounded ~10k labeling sample, not the corpus).
- **2026-05-25 (ops)** — Recorded operational policy from prior-run experience: per-batch
  scoring checkpoints (≤30-min loss target; implemented in 06c), ~4 parallel GPU scoring workers
  on the Blackwell (sharded by year; benchmark N), DiD is fast (revised that row down ~4–8 h →
  ~1–3 h), and the usable-year count is ~172 (not 180 files); analysis only needs 1895–1945.
- **2026-05-26 (overnight-run #1 actuals + clean bottleneck found)** — First unattended run
  (`run_overnight.ps1`, ~14.5 h wall-clock) **never reached the analysis window.** The CLEAN
  stage (02) was **serial and unscoped**, so it processed years from 1770 forward and reached
  only ~1770–1869 (~100 partitions) before the night ended — lexicon/features/RF for 1895–1945
  never ran. Root cause: serial clean is far slower than the ~2–4 h estimate on the heavy
  mid-corpus years (e.g., 1856 = 1.43M articles → 1.29M kept, ~90%; tens of min each). **Fixes:**
  (a) parallelized clean across years with a PSOCK cluster (cap ~10 workers; RAM-bound), atomic
  per-year writes, skip-if-done — in `rmd/02_clean_articles.Rmd`; (b) scope clean to the window
  via `CAPLAN_CLEAN_YEARS=1895–1945` in `run_overnight.ps1` so it stops wasting time on
  out-of-window years. The ~1770–1869 partitions already cleaned are moved to
  `articles_cleaned_prewindow/` (preserved, not deleted) so the window dataset is exactly
  1895–1945 for downstream 06a/06b/06c with no code changes. Reboots during the run were mostly
  user-initiated; the script is resumable, so each relaunch skips completed work. **Lesson:**
  scope *every* stage to 1895–1945, not just scoring; the clean estimate row is updated in §2.
- **2026-05-26 (scope REVERSAL: measure the WHOLE corpus; window is DiD-only)** — Supersedes the
  prior bullet's "scope every stage to 1895–1945" lesson, per user directive: **score everything as
  if the DiD step didn't exist.** Rationale: (a) more pre-treatment years *strengthen* the DiD
  (longer parallel-trends / event-study leads), so out-of-window years are useful signal, not waste;
  (b) the DiD (09) is a tiny step relative to cleaning + scoring the corpus, so it must not constrain
  the expensive measurement stages. `YEARS_ANALYSIS` (1895–1945) is demoted to its real role — the
  **shock window for stages 08/09 only** — and decoupled from measurement. Config: `CAPLAN_CLEAN_YEARS`
  and now `CAPLAN_SCORE_YEARS` are both the **full corpus, window-first** (focal years land first;
  ordering, not a limit). The earlier plan to move pre-window partitions to `articles_cleaned_prewindow/`
  is **dropped** — all years stay in `articles_cleaned`. **To-do (attended LLM step):** widen
  `04_sample_for_labeling` and the RoBERTa scorer (06) from `YEARS_ANALYSIS` to the full corpus so the
  training set + classifier span every era. **Runtime:** full-corpus scoring is ~3× the windowed load
  — the timeline long pole (weeks–months class, §3), accepted as the planned long-running hand-off;
  $0 API, resumable, window-first so analysis-ready years come first.
